Inferring the effective start dates of non-pharmaceutical interventions during \covid outbreaks

### 2021-01-22 ms/ms.tex
typos fix

We compare the posterior predictive plots of a model with  free±$\tau$ with those of a model with~$\tau$ fixed at~$\tau^*$ and a model without~$\tau$ (i.e. transmission and reporting rates are constant).
All three models were fitted to case data up to Apr~11, used to predict out-of-sample case data up to Apr~24, and these predictions were then compared to the real case data. 
The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}):
In all 11 european countries, the expected posterior RMSE (root mean squared error) of the out-of-sample predictions is lowest for the model with a free~$\tau$ (\autoref{table:RMSE}).
When we compare the models using WAIC~(\autoref{eq:WAIC}), the model with free~$\tau$ is strongly preferred in 9 out of 12 countries, the exceptions being Norway (only marginal preference), Sweden, and Denmark (\autoref{table:WAIC}).
Indeed, we estimate low effect of NPIs on transmission in Denmark and Sweden (i.e. $\lambda=0.7$ and $0.74$, respectively; see~\autoref{table:estimated-params}).
This may interfere with the inference of $\tau$ due to unidentifiability. 
Notably, the data for Sweden and Denmark do not have a single "peak" during the evaluated dates, possibly leading to wide credible intervals on $\tau$ (\autoref{fig:tau-summary}) and poor WAIC in the model with free~$\tau$, whereas the duration between the first and last interventions was especially long in Norway (\autoref{table:NPI_dates}).

@@

In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (final score 2--1) with a crowd of 50,965 in Seville.
A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
Nevertheless, we estimate the effective start of NPI eleven days later, at Mar~25($\substack{+1.70 \\ -1.43}$~days 95 \%CI, \autoref{fig:late}).

@@

We also estimate an early effective start of NPIs in Denmark, Norway, and Sweden.
However, the credible intervals are quite wide (\autoref{fig:tau-summary}), and in Denmark and Sweden the evidence did not support the model with free±$\tau$ over the model with $\tau$ fixed at the official date (WAIC values in \autoref{table:WAIC}).
Indeed, Denmark and Sweden had the smallest estimated effect of NPIs on transmission rates, which  probably hinders our ability to estimate $\tau$ in these countries (\autoref{table:estimated-params}).
Moreover, in Sweden the number of daily cases continued to grow up to Apr~11, rather than "peak" (\autoref{fig:ppc}). 
In Denmark, the opposite occurred: there were seemingly two "peaks", on Mar~11 and on Apr~8 (\autoref{fig:ppc}); the first "peak" may be a result of stochastic events, for example due to a large cluster of cases or an accumulation of tests.
We suspect that these missing and additional "peaks" increase the uncertainty in our inference.

@@

    Circles and stars represent daily case data up to and after Apr~11, respectively

>>>>

We compare the posterior predictive plots of a model with  free~$\tau$ with those of a model with~$\tau$ fixed at~$\tau^*$ and a model without~$\tau$ (i.e. transmission and reporting rates are constant).
All three models were fitted to case data up to Apr~11, used to predict out-of-sample case data up to Apr~24, and these predictions were then compared to the real case data. 
The model with free~$\tau$ clearly produces better and less variable predictions~(\autoref{fig:ppc}):
In all 11 european countries, the expected posterior RMSE (root mean squared error) of the out-of-sample predictions is lowest for the model with a free~$\tau$ (\autoref{table:RMSE}).
When we compare the models using WAIC~(\autoref{eq:WAIC}), the model with free~$\tau$ is strongly preferred in 9 out of 12 countries, the exceptions being Norway (only marginal preference), Sweden, and Denmark (\autoref{table:WAIC}).
Indeed, we estimate low effect of NPIs on transmission in Denmark and Sweden (i.e. $\lambda=0.7$ and $0.74$, respectively; see~\autoref{table:estimated-params}).
This may interfere with the inference of $\tau$ due to unidentifiability. 
Notably, the data for Sweden and Denmark do not have a single "peak" during the evaluated dates, possibly leading to wide credible intervals on $\tau$ (\autoref{fig:tau-summary}) and poor WAIC in the model with free~$\tau$, whereas the duration between the first and last interventions was especially long in Norway (\autoref{table:NPI_dates}).

@@

In Spain, social distancing was encouraged starting on Mar~8~\citep{Flaxman2020}, but mass gatherings still occurred on Mar~8, including a march of 120,000 people for the \href{https://www.nytimes.com/2020/03/13/world/europe/spain-coronavirus-emergency.html}{International Women's Day}, and a football match between \href{https://www.espn.com/soccer/match?gameId=550350}{Real Betis and Real Madrid} (final score 2--1) with a crowd of 50,965 in Seville.
A national lockdown was only announced on Mar~14~\citep{Flaxman2020}.
Nevertheless, we estimate the effective start of NPI eleven days later, at Mar~25($\substack{+1.70 \\ -1.43}$~days 95\% CI, \autoref{fig:late}).

@@

We also estimate an early effective start of NPIs in Denmark, Norway, and Sweden.
However, the credible intervals are quite wide (\autoref{fig:tau-summary}), and in Denmark and Sweden the evidence did not support the model with free~$\tau$ over the model with $\tau$ fixed at the official date (WAIC values in \autoref{table:WAIC}).
Indeed, Denmark and Sweden had the smallest estimated effect of NPIs on transmission rates, which  probably hinders our ability to estimate $\tau$ in these countries (\autoref{table:estimated-params}).
Moreover, in Sweden the number of daily cases continued to grow up to Apr~11, rather than "peak" (\autoref{fig:ppc}). 
In Denmark, the opposite occurred: there were seemingly two "peaks", on Mar~11 and on Apr~8 (\autoref{fig:ppc}); the first "peak" may be a result of stochastic events, for example due to a large cluster of cases or an accumulation of tests.
We suspect that these missing and additional "peaks" increase the uncertainty in our inference.

@@

    Circles and stars represent daily case data up to and after Apr~11, respectively.
